Supplementary Information:\\Evidence for black-swan events in animal populations

### 2015-11-24 text/supp.tex
Update journal format

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose, letterpaper, tmargin = 2.54cm, bmargin = 2.54cm,
  lmargin = 2.54cm, rmargin = 2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{ragged2e}
\usepackage{verbatim}
\usepackage[sectionbib]{chapterbib}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{tabu} % for spacing between rows in longtable
\usepackage{titlesec}
\usepackage{url}
\usepackage{mathtools}
\usepackage{cite}
\usepackage{citesupernumber}

@@

% Linux Libertine:
\usepackage{textcomp}
\usepackage[sb]{libertine}
\usepackage[varqu,varl]{inconsolata}% sans serif typewriter
\usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
\usepackage[cal=boondoxo]{mathalfa} % mathcal
\useosf % osf for text, not math
\usepackage[supstfm=libertinesups,%
  supscaled=1.2,%
  raised=-.13em]{superiors}

@@

\usepackage[round,sectionbib]{natbib}
\bibpunct{(}{)}{;}{a}{}{;}

@@

@@

@@

\item We removed all populations rated as $< 2$ in the GPDD quality assessment
  (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
  \citep[following][]{sibly2005, ziebarth2010}.

\item Populations with negative abundance values were removed. Of the
  populations that remained at the end of our other filtering rules, the
  remaining populations with negative abundances listed were all from time
  series that had been standardized by subtracting the mean and dividing by the
  standard deviation. We verified this by locating the original papers the
  datasets were extracted from: \citet{colebrook1978} for zooplankton and
  \citet{lindstrom1995} for grouse. Since the papers did not include the
  original mean time-series values we could not back transform these data
  points.

@@

\item We filled in single recorded values of zero with the lowest non-zero
  value in the time series \citep[following][]{brook2006a}. This assumes that
  single values of zero result from abundance being low enough that censusing
  overlooked individuals that were actually present. We turned multiple zero
  values in a row into \texttt{NA} values. This implies that multiple zero
  values were either censusing errors or caused by emigration. Regardless, our
  population models were fit on a multiplicative (log) scale and so could not
  account for zero abundance. To avoid distorting the original data too
  strongly, we removed populations in which we filled in more than four zeros.

\item We removed all populations without at least four unique values
  \citep[following][]{brook2006a}.

@@

\item We then wrote an algorithm to find the longest unbroken window of
  abundance (no \texttt{NA}s) with at least $20$ time steps in each population
  time series. If there were any populations with multiple windows of identical
  length, we took the most recent window. This is a longer window than used in
  some previous analyses \citep[e.g.][]{brook2006a}, but since our model
  attempts to capture the shape of the distribution tails, our model requires
  more data.

@@

\item Finally, we removed populations with the following GPDD Main IDs, which
  we discovered were data entry errors when verifying the populations with
  suspected black swans: \texttt{1207} because the 1957 data point was entered
  as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because the
  1978 data point was entered as 7 but should have been 47 \citep{minot1986},
  and \texttt{6566} because some of the data did not match the graph
  \citep{heessen1996}.

@@

The probability density for the skewed Student-t distribution \citep{fernandez1998} equates to the following probability density as implemented in \citet{king2012}:

Since Stan only needs the probability density up to an additive
constant \citep{stan-manual2014}, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student's t distribution implemented in
Stan\footnote{Credit to Stan developer Bob Carpenter:
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

@@

We then fit our Gompertz models to the simulated datasets with all parameters
(except $\nu$) set near the median values estimated in the GPDD. We repeated
this with $50$ and $100$ samples without observation error, $50$ samples with
observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the
same observation error and a Gompertz model that allowed for correctly
specified observation error magnitude. Our results indicate that the Gompertz
model can recapture the true value of $\nu$ when the process noise was chosen
so that appropriate tail events were present (Extended Data Fig.
5 upper panels). The addition of observation error caused the
model to tend to underestimate the degree of heavy-tailedness. Fitting a model
with correctly specified observation error did not make substantial
improvements to model bias (Extended Data Fig. 5).

When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
models distinguished heavy and not-heavy tails reasonably well (Extended Data
Fig. 5 lower panels). Without observation error, and using a
probability of $0.5$ as a threshold, the model correctly classified all
simulated systems with normally distributed process noise as not heavy tailed.
The model would have miscategorized only one of $40$ simulations at $\nu = 5$
across simulated populations with $50$ or $100$ time steps (Extended Data Fig.
5, scenarios 1 and 2 in lower row, second panel from left).
The model would have correctly categorized all cases where the process noise
was not heavy tailed (Extended Data Fig. 5 bottom-right
panel) and all cases where $\nu = 3$ and there was not observation error. With
$0.2$ standard deviations of observation error, the model still categorized
\obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one
case when $\nu = 3$. Allowing for observation error made little improvement to
the detection of heavy tails. Therefore, we chose to focus on the simpler
model without observation error in the main text, particularly given that the
true magnitude of observation error was unknown in the empirical data.

@@

\bibliographystyle{apalike}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,supp}
%
% ------------------------------
% Supplemental Tables
% ------------------------------

>>>>

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose, letterpaper, tmargin = 2.54cm, bmargin = 2.54cm,
  lmargin = 2.54cm, rmargin = 2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{ragged2e}
\usepackage{verbatim}
\usepackage[sectionbib]{chapterbib}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{tabu} % for spacing between rows in longtable
\usepackage{titlesec}
\usepackage{url}
\usepackage{mathtools}
%\usepackage{cite}
\usepackage{citesupernumber}
\usepackage{scicite}

@@

% Linux Libertine:
\usepackage{textcomp}
\usepackage[sb]{libertine}
\usepackage[varqu,varl]{inconsolata}% sans serif typewriter
\usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
\usepackage[cal=boondoxo]{mathalfa} % mathcal
%\useosf % osf for text, not math
\usepackage[supstfm=libertinesups,%
  supscaled=1.2,%
  raised=-.13em]{superiors}

@@

%\usepackage[round,sectionbib]{natbib}
%\bibpunct{(}{)}{;}{a}{}{;}

@@

\nocite{taleb2007,sornette2009}
\nocite{meehl2004,katz2005,ipcc2012}
\nocite{taleb2007,sornette2009,may2008}
\nocite{taleb2007}
\nocite{newman2005}
\nocite{sornette2009}
\nocite{harnik2012}
\nocite{fey2015}
\nocite{mangel1994}
\nocite{ipcc2012}
\nocite{keitt1998,allen2001,halley2002}
\nocite{brook2006a,knape2012}
\nocite{SOM}
\nocite{saucy1994}
\nocite{stafford1971}
\nocite{stafford1971}
\nocite{potts1980}
\nocite{potts1980}
\nocite{allen2001}
\nocite{schindler2010}
\nocite{carpenter2006}
\nocite{nunez2012}
\nocite{meehl2004,katz2005,ipcc2012}
\nocite{kirby2009}
\nocite{denny2009}
\nocite{nrc2007}
\nocite{caddy1996}
\nocite{schindler2010}
\nocite{doak2008,lindenmayer2010}
\nocite{mangel1994}
\nocite{loreau2010a,thompson2013}
\nocite{meehl2004,ipcc2012,thompson2013}
\nocite{rachev2008}

@@

We selected abundance time series from the Global Population Dynamics Database\cite{gpdd2010} (GPDD), which contains nearly 5,000 time series of abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We filtered the data (Supplementary Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}). Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (table~S1) and none of the data points that were later considered black-swan events. Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range \minTimeSteps--\maxTimeSteps) (table~S1).

@@

\item We removed all populations rated as $< 2$ in the GPDD quality assessment
  (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
  \cite{sibly2005, ziebarth2010}.

\item Populations with negative abundance values were removed. Of the
  populations that remained at the end of our other filtering rules, the
  remaining populations with negative abundances listed were all from time
  series that had been standardized by subtracting the mean and dividing by the
  standard deviation. We verified this by locating the original papers the
  datasets were extracted from: \cite{colebrook1978} for zooplankton and
  \cite{lindstrom1995} for grouse. Since the papers did not include the
  original mean time-series values we could not back transform these data
  points.

@@

\item We filled in single recorded values of zero with the lowest non-zero
  value in the time series \cite{brook2006a}. This assumes that
  single values of zero result from abundance being low enough that censusing
  overlooked individuals that were actually present. We turned multiple zero
  values in a row into \texttt{NA} values. This implies that multiple zero
  values were either censusing errors or caused by emigration. Regardless, our
  population models were fit on a multiplicative (log) scale and so could not
  account for zero abundance. To avoid distorting the original data too
  strongly, we removed populations in which we filled in more than four zeros.

\item We removed all populations without at least four unique values
  \cite{brook2006a}.

@@

\item We then wrote an algorithm to find the longest unbroken window of
  abundance (no \texttt{NA}s) with at least $20$ time steps in each population
  time series. If there were any populations with multiple windows of identical
  length, we took the most recent window. This is a longer window than used in
  some previous analyses \cite{brook2006a}, but since our model
  attempts to capture the shape of the distribution tails, our model requires
  more data.

@@

\item Finally, we removed populations with the following GPDD Main IDs, which
  we discovered were data entry errors when verifying the populations with
  suspected black swans: \texttt{1207} because the 1957 data point was entered
  as 2 but should have been 27 \cite{kendeigh1982}, \texttt{6531} because the
  1978 data point was entered as 7 but should have been 47 \cite{minot1986},
  and \texttt{6566} because some of the data did not match the graph
  \cite{heessen1996}.

@@

\section{Details on the heavy-tailed Gompertz probability model} We fit heavy-tailed Gompertz population dynamics models to data from the GPDD. The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
\end{align}
The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\). The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1a, b).

For the Gompertz model, our weakly-informative priors (fig.~\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We would not expect values of \(b\) with levels of density dependence as low as \(-1\) (very strong inverse density dependence), nor would we generally expect values above \(1\). We allowed values of \(b\) above \(1\) to allow for non-stationary time series of growth rates. The estimates of \(b\) were well within these bounds. Our prior on \(\lambda\) was weakly informative within the range of expected values for population growth and is similar to the default priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the specific scale parameter of \(2.5\) is based on our expected range of the value in nature from previous studies\cite{connors2014}. In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to a uniform prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}. We truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid drifting towards infinity. In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1). Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics. When categorizing a population as heavy tailed, we applied a 0.5 probability threshold. In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior for \(\nu\) will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the population as heavy tailed.

We fit our models with Stan 2.6.0\cite{stan-manual2015, hoffman2014}, and R 3.2.0\cite{r2015}. We began with four chains and 2,000 iterations, discarding the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the potential scale reduction factor---a measure of chain convergence) was greater than \(1.05\) for any parameter, or if the minimum effective sample size, \(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples), for any parameter was less than \(200\), we doubled both the total iterations and warm up period and sampled from the model again. These thresholds are in excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of \(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and confidence intervals. In the majority of cases our minimum thresholds were greatly exceeded. We continued this procedure up to 8,000 iterations (16,000 total samples) by which all chains were deemed to have sufficiently converged. The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally requires fewer iterations to obtain equivalent effective sample sizes than the Gibbs or Metropolis--Hastings algorithms\cite{stan-manual2015}.

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ. Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supplementary Information). Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

\section{Alternative priors} To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors. Our base model used a prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (fig.~\ref{fig:priors}c). Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\)) prior on \(1/\nu\) as used in Gelman \emph{et al.}\cite{gelman2014}, but this prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\) than the truncated exponential priors considered here (not shown).

Our results show that weaker and stronger priors have little influence on our conclusions about heavy-tailed dynamics (fig.~\ref{fig:alt-priors}). When the data are informative about tail behaviour (i.e. when there is strong evidence of low \(\nu\) values, upper-right of fig.~\ref{fig:alt-priors}), the prior has little impact on the estimate of \(\nu\). When the data are less informative about \(\nu\) (i.e.\ when there are no or few tail events and time series are short or noisy), the prior can pull the estimate of \(\nu\) towards larger or smaller values (fig.~\ref{fig:alt-priors}). The vast majority of the populations with Pr\((\nu < 10)\) in the base prior were not altered qualitatively by this range of prior strength.

\section{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions. The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.

\subsection{Autocorrelated residuals} We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process-noise residuals:
\begin{align}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align}
In addition to the parameters in the original Gompertz model, this model estimates an additional parameter \(\phi\), which represents the correlation of subsequent residuals. Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen prior in previous analyses\cite{thorson2014a}, we placed a moderately informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)\).

\subsection{Assumed density independence} We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent). This is equivalent to fitting a random walk model (with drift) to the \(\ln\) abundances or assuming the growth rates are drawn from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of \(\nu\), (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence\cite{segura2013}.

\subsection{Assumed observation error} Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}. In our main analysis, we fit a model that ignored observation error. One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error. However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model\cite{knape2008}. Furthermore, our model was applied to hundreds of time series, often of short length (as few as \minTimeSteps\ time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater. Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\) represents the standard deviation of observation error (on a log scale), and \(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set \(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in simulation analyses\cite{valpine2002, thorson2014b}.

\subsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance. In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance.

To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance (similar and less informative than in Delean \emph{et al.}\cite{delean2013}). We set the prior on \(r_\mathrm{max}\) as uniform between 0 and 20 as in Delean \emph{et al.}\cite{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in the Gompertz model.

\section{Simulation testing} We performed two types of simulation testing (Supporting Information). First, we tested how easily the Student-t distribution \(\nu\) parameter could be recovered given different true values of \(\nu\) and different sample sizes. Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of \(\nu\) given that a set of process deviations was provided in which the effective \(\nu\) value was close to the true \(\nu\) value.

We found that we could consistently and precisely recover median posterior estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge 200\)) (fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were. At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (fig.~\ref{fig:sim-nu} lower panels).

Our simulation-testing results indicate that the Gompertz model can recapture the true value of \(\nu\) when the process noise was chosen so that appropriate tail events were present (fig.~\ref{fig:sim-prob} upper panels). The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness. Fitting a model with correctly specified observation error did not make substantial improvements to model bias (fig.~\ref{fig:sim-prob}).

@@

We then fit our Gompertz models to the simulated datasets with all parameters
(except $\nu$) set near the median values estimated in the GPDD. We repeated
this with $50$ and $100$ samples without observation error, $50$ samples with
observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the
same observation error and a Gompertz model that allowed for correctly
specified observation error magnitude. Our results indicate that the Gompertz
model can recapture the true value of $\nu$ when the process noise was chosen
so that appropriate tail events were present (fig.
\ref{fig:sim-prob} upper panels). The addition of observation error caused the
model to tend to underestimate the degree of heavy-tailedness. Fitting a model
with correctly specified observation error did not make substantial
improvements to model bias (fig. \ref{fig:sim-prob}).

When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well (fig. \ref{fig:sim-prob} lower panels). Without observation error, and using a probability of $0.5$ as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed.  The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across simulated populations with $50$ or $100$ time steps (fig. \ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left).  The model would have correctly categorized all cases where the process noise was not heavy tailed (fig. \ref{fig:sim-prob} bottom-right panel) and all cases where $\nu = 3$ and there was not observation error. With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one case when $\nu = 3$. Allowing for observation error made little improvement to the detection of heavy tails. Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.

\section{Covariates of heavy-tailed dynamics} We fit a hierarchical beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics. We obtained maximum lifespan and body-size data from Brook \emph{et al.}\cite{brook2006a}. The beta distribution is useful when response data are on a continuous scale between zero and one\cite{ferrari2004}. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients. The intercepts are allowed to vary from the overall intercept \(\alpha\) by taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order (\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \; \mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters \(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to match the way they are visually represented in fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale. All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude\cite{gelman2008c}. We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length. Lifespan is also more directly related to time and potential mechanisms driving black-swan dynamics.

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters\cite{gelman2006c, gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior. We fit our models with 5,000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample. We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters). We further checked our model against a similar model with the glmmADMB \textsf{R} package\cite{glmmadmb} without uncertainty around the predictors.

To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate hierarchical model with the same structure but with only \(\log\) time-series length as a predictor---in this case, we did not want to control for intrinsic population characteristics such as density dependence. Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).

\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise\cite{fernandez1998} (Supplementary Information). The distribution adds one parameter to the Student-t distribution, \(\gamma\), which controls the skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0 < \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We aggregated 5,000 randomly selected posterior samples from the \(\gamma\) parameter of each model at three levels of evidence for heavy-tails: \emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le \mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\) (Fig.~4).

To generate five-year forecasts of abundance, we combined the posterior parameter samples from the skewed-t Gompertz models with stochastically generated process noise. We compared these forecasts to those generated by a standard Gompertz model with normally distributed process noise fit to the same data. We calculated the ratio of the lower 99\% quantile credible interval between the two model projections for all populations in which Pr(\(\nu < 10\)) \(> 0.5\). To ensure the results had stabilized in the tails of the forecast posterior, we increased the number of posterior samples in Stan. We ran four chains of 20,000 iterations and discarded the first 10,000 as warmup for a total of 40,000 samples.

The probability density for the skewed Student-t distribution \cite{fernandez1998} equates to the following probability density as implemented in Ref.~\cite{king2012}:

Since Stan only needs the probability density up to an additive
constant \cite{stan-manual2014}, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student's t distribution implemented in
Stan\footnote{Credit to Stan developer Bob Carpenter:
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

@@

\section{Code availability.} Data and code to reproduce our analysis are available at \url{https://github.com/seananderson/heavy-tails} (currently private).

\bibliographystyle{Science}
%\bibliography{/Users/seananderson/Dropbox/tex/jshort,supp}
\bibliography{jshort,newbib}

\section{Supplementary tables and figures}

%\newenvironment{helvetica}{\fontfamily{phv}\selectfont}{\par}

\singlespacing

\textbf{Table~S1: Summary statistics for the filtered Global Population Dynamics Database time series arranged by taxonomic class.} Columns are number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.

\onehalfspacing

%\begin{helvetica}
\smallskip
\begin{scriptsize}
\begin{tabular}{lrrrrrrrr}
\toprule
\input{../analysis/stat-table.tex}
\label{tab:stats}
\end{tabular}
\end{scriptsize}
%\end{helvetica}

\LTcapwidth=\textwidth

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}

\renewcommand{\tablename}{Table}

\setcounter{table}{1}

\renewcommand{\arraystretch}{0.1}% Tighter

%\begin{helvetica}
\singlespacing
\begin{scriptsize}
\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
\caption{\textbf{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.} Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses. Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted t distribution.}\\

\toprule
\input{../analysis/cause-table.tex}
\label{tab:causes-supp}
\end{longtable}
\end{scriptsize}
\onehalfspacing
%\end{helvetica}

\renewcommand{\figurename}{Figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{../analysis/priors-gomp-base.pdf}
